---
title: "Complexity Engine: Early Results Reference"
author: "Ali M. Al-Timimi"
date: "2026-05-09"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
  pdf:
    toc: true
execute:
  echo: true
  warning: true
  message: true
---

# Purpose

This document is a working reference for the first successful run of the complexity structural engine within the Global Cancer Complexity structural inference framework.

The goal is not to provide a final interpretation, but to preserve the current code, diagnostic plots, and early observations in a reproducible form.

At this stage, the complexity engine has successfully propagated the following identifiers across the structural inference pipeline:

- `chip`
- `group`
- `comparison`
- `filter_regime`
- `engine`
- `gene_set_mode`
- `gene_set_name`
- complexity metrics
- tumor-minus-normal delta metrics

The current implementation focuses on `gene_set_mode = "FULL"`, meaning complexity is computed over the full filtered probe set for each normal--tumor comparison and probe-selection regime.

# Conceptual framing

The complexity engine treats each normal–tumor comparison as a transformation in the structural organization of a filtered expression matrix.

For each comparison, the engine resolves the paired normal and tumor matrices, restricts both matrices to the selected probes, and computes SVD-based and related matrix-structural complexity summaries.

The current focus is on whether tumors show evidence of:

- increased or decreased SVD-based condition structure;
- changes in effective rank;
- changes in matrix sparsity;
- changes in composite condition-number structure;
- consistency across chip platforms and probe-selection regimes.

# Load results

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

source(here::here("R/config/global_cancer/structural_inference_config.R"), local = FALSE)

complexity_results <- readRDS(
  file.path(structural_rdata_dir, "complexity_results.rds")
)
```

# Inspect object structure

```{r}
names(complexity_results)
str(complexity_results, max.level = 1)
```

Expected structure:

```text
summary: paired normal--tumor complexity summaries
```

# Inspect summary table

```{r}
complexity_df <- complexity_results$summary

dplyr::glimpse(complexity_df)
```

Check computation status:

```{r}
complexity_df |>
  dplyr::count(status)
```

Check that chip and regime propagation worked:

```{r}
complexity_df |>
  dplyr::count(chip, filter_regime)
```

Check unique probe-selection regimes:

```{r}
unique(complexity_df$filter_regime)
```

Expected regimes from current configuration:

```text
limma
variance_top3000
```

## Early interpretation

Successful recovery of both `limma` and `variance_top3000` confirms that the complexity engine now follows the same regime-aware structural architecture as the MP spectral engine.

This means the following chain is functioning:

```text
config
→ pipeline
→ wrapper
→ pairwise complexity engine
→ paired comparison layer
→ core complexity metrics
→ standardized summary table
```

The initial run produced a small number of expected skips for limma comparisons with too few retained probes. These should be retained as QC information rather than treated as software failures.

# Current complexity metrics

The main paired complexity output currently includes:

- `kappa_normal`
- `kappa_tumor`
- `kappa_delta`
- `direction`
- `p_perm`
- `p_perm_fdr`
- bootstrap confidence intervals for normal and tumor complexity
- `effrank_normal`
- `effrank_tumor`
- `effrank_delta`
- `sparsity_normal`
- `sparsity_tumor`
- `sparsity_delta`
- `composite_kappa_normal`
- `composite_kappa_tumor`
- `composite_kappa_delta`

## Statistical notes

These should be interpreted cautiously as structural descriptors rather
than as finalized biological claims. Importantly, these metrics may
capture partially distinct structural dimensions rather than redundant
summaries of a single underlying quantity.

The permutation p-values (`p_perm`) are currently computed using a
two-sided permutation framework:

mean(abs(null_distribution) >= abs(observed_difference))

This choice reflects the current exploratory nature of the structural
inference framework, where both complexity gain and complexity loss are
considered biologically meaningful possibilities.

Adjusted permutation p-values (`p_perm_fdr`) are computed using the
Benjamini–Hochberg procedure within each
`chip × filter_regime × gene_set_mode` stratum.

Bootstrap and permutation procedures are seeded for reproducibility.

# Plot 1: Complexity delta by chip and regime

```{r}
ggplot(
  complexity_df,
  aes(
    x = filter_regime,
    y = kappa_delta
  )
) +
  geom_boxplot() +
  facet_wrap(~ chip) +
  labs(
    title = "SVD kappa delta by chip and probe-selection regime",
    x = "Filter regime",
    y = "Tumor-minus-normal SVD kappa delta"
  )
```

## Early interpretation

This plot provides a compact regime-level view of complexity gain or loss.

Positive `kappa_delta` indicates that tumor matrices have higher SVD-based condition structure than their matched normal matrices. Negative `kappa_delta` indicates reduced SVD-based condition structure in tumor.

The interpretation should remain cautious because condition-number-based metrics can be sensitive to feature count, sample count, and matrix rank properties.

# Plot 2: Effective rank delta by chip and regime

```{r}
ggplot(
  complexity_df,
  aes(
    x = filter_regime,
    y = effrank_delta
  )
) +
  geom_boxplot() +
  facet_wrap(~ chip) +
  labs(
    title = "Effective rank delta by chip and probe-selection regime",
    x = "Filter regime",
    y = "Tumor-minus-normal effective rank delta"
  )
```

## Early interpretation

Effective rank provides a complementary measure of dimensional participation.

Positive `effrank_delta` suggests that tumor-associated variation is distributed across more effective dimensions. Negative `effrank_delta` suggests a collapse toward fewer effective dimensions.

This metric should later be compared directly with MP participation-ratio and spectral-entropy deltas.

# Plot 3: SVD kappa delta vs effective rank delta

```{r}
ggplot(
  complexity_df,
  aes(
    x = effrank_delta,
    y = kappa_delta,
    color = filter_regime
  )
) +
  geom_point(size = 3) +
  facet_wrap(~ chip) +
  labs(
    title = "SVD kappa delta vs effective rank delta",
    x = "Effective rank delta",
    y = "SVD kappa delta",
    color = "Filter regime"
  )
```

## Early interpretation

This plot begins to test whether complexity gain or loss is associated with changes in effective dimensionality.

Several patterns are possible:

| Pattern | Possible interpretation |
|---|---|
| `kappa_delta > 0`, `effrank_delta > 0` | Increased condition structure with broader dimensional participation. |
| `kappa_delta > 0`, `effrank_delta < 0` | Increased condition structure concentrated in fewer dimensions. |
| `kappa_delta < 0`, `effrank_delta > 0` | Reduced condition-number structure but broader dimensional distribution. |
| `kappa_delta < 0`, `effrank_delta < 0` | Reduced complexity and reduced effective dimensionality. |

# Plot 4: Composite kappa delta vs SVD kappa delta

```{r}
ggplot(
  complexity_df,
  aes(
    x = kappa_delta,
    y = composite_kappa_delta,
    color = filter_regime
  )
) +
  geom_point(size = 3) +
  facet_wrap(~ chip) +
  labs(
    title = "Composite kappa delta vs SVD kappa delta",
    x = "SVD kappa delta",
    y = "Composite kappa delta",
    color = "Filter regime"
  )
```

## Early interpretation

This diagnostic checks whether the primary SVD kappa metric is aligned with the broader composite condition-number summary.

Strong agreement would suggest that complexity changes are not specific to one condition-number estimator. Disagreement would suggest that different aspects of matrix conditioning are being captured by different metrics.

# Plot 5: Complexity direction counts

```{r}
complexity_df |>
  dplyr::count(chip, filter_regime, direction) |>
  ggplot(
    aes(
      x = filter_regime,
      y = n,
      fill = direction
    )
  ) +
  geom_col(position = "dodge") +
  facet_wrap(~ chip) +
  labs(
    title = "Complexity gain/loss counts by chip and regime",
    x = "Filter regime",
    y = "Number of comparisons",
    fill = "Direction"
  )
```

## Early interpretation

This plot provides a coarse summary of how often complexity is gained or lost under each chip and probe-selection regime.

It should not be treated as a final biological result, but it is useful for detecting gross asymmetry across platforms or filtering regimes.

# Correlation analysis of complexity metrics

```{r}
complexity_metrics <- complexity_df |>
  dplyr::select(
    kappa_delta,
    effrank_delta,
    sparsity_delta,
    composite_kappa_delta
  )

cor(complexity_metrics, use = "pairwise.complete.obs")
```

## Early interpretation

This correlation matrix is intended to identify whether the complexity metrics are capturing a shared structural axis or multiple distinct axes. The current results suggest that complexity metrics are not fully reducible to a single latent structural dimension.

In particular, `kappa_delta` and `composite_kappa_delta` are strongly aligned, whereas `effrank_delta` appears comparatively orthogonal to the condition-number metrics. This suggests that condition-number complexity and effective-dimensional complexity may represent partially distinct forms of structural organization.
Potential relationships to inspect:

- `kappa_delta` vs `composite_kappa_delta`;
- `kappa_delta` vs `effrank_delta`;
- `effrank_delta` vs `sparsity_delta`;
- `composite_kappa_delta` vs `effrank_delta`.

The key question is whether condition-number complexity and effective-dimensionality complexity move together or diverge across tumor transitions.

For the current FULL filtered-probe runs, `sparsity_delta` was constant across comparisons and therefore contributed no variance to the correlation or PCA analyses. This metric is retained in the engine output for future regimes and gene-set modes, but is currently non-informative for the present dataset configuration.

# PCA of complexity delta metrics

```{r}
complexity_metrics <- complexity_df |>
  dplyr::select(
    kappa_delta,
    effrank_delta,
    sparsity_delta,
    composite_kappa_delta
  )

complexity_metrics_clean <- complexity_metrics |>
  tidyr::drop_na()

nonconstant_cols <- complexity_metrics_clean |>
  dplyr::summarise(
    dplyr::across(
      dplyr::everything(),
      ~ sd(.x, na.rm = TRUE) > 0
    )
  ) |>
  unlist()

complexity_metrics_pca <- complexity_metrics_clean[, nonconstant_cols, drop = FALSE]

complexity_pca <- prcomp(
  complexity_metrics_pca,
  scale. = TRUE
)

summary(complexity_pca)
round(complexity_pca$rotation, 3)
```

## PCA score plot

```{r}
pca_row_idx <- stats::complete.cases(complexity_metrics) &
  apply(complexity_metrics, 1, function(x) all(is.finite(x)))

complexity_pca_plot_df <- complexity_df[pca_row_idx, ] |>
  dplyr::select(chip, group, comparison, filter_regime) |>
  dplyr::bind_cols(as.data.frame(complexity_pca$x))

ggplot(
  complexity_pca_plot_df,
  aes(
    x = PC1,
    y = PC2,
    color = filter_regime,
    shape = chip
  )
) +
  geom_point(size = 3) +
  labs(
    title = "PCA of complexity delta metrics",
    x = "PC1",
    y = "PC2",
    color = "Filter regime",
    shape = "Chip"
  )
```

The PCA decomposition further supports a multi-axis interpretation of transcriptomic structural complexity.

PC1 is dominated by condition-number-related metrics:

- `kappa_delta`
- `composite_kappa_delta`

whereas PC2 is dominated almost entirely by:

- `effrank_delta`

This suggests that effective dimensionality behaves semi-independently from condition-number structure within the current FULL filtered-probe framework.

Accordingly, tumor-associated structural change may involve:

- increased condition structure with reduced dimensional participation;
- reduced condition structure with expanded dimensional participation;
- or coordinated shifts in both properties.

This argues against interpreting “complexity” as a single scalar biological property.

## Early interpretation

The PCA should be used to determine whether complexity metrics collapse onto a dominant structural axis, as was observed in the MP spectral engine, or whether complexity decomposes into several partially independent axes.

Possible axes include:

- condition-number amplification;
- effective-dimensionality shift;
- sparsity or near-zero structure;
- composite matrix conditioning.

# Cross-engine alignment with MP results

Once both MP and complexity outputs are available, a useful next diagnostic is to join complexity deltas with MP spectral deltas by:

```text
chip
group
comparison
filter_regime
```

```{r}
mp_spectral_results <- readRDS(
  file.path(structural_rdata_dir, "mp_spectral_results.rds")
)

mp_delta_df <- mp_spectral_results$deltas

combined_df <- complexity_df |>
  dplyr::inner_join(
    mp_delta_df,
    by = c("chip", "group", "comparison", "filter_regime"),
    suffix = c("_complexity", "_mp")
  )

dplyr::glimpse(combined_df)
```

## Plot: Complexity delta vs MP participation-ratio delta

```{r}
ggplot(
  combined_df,
  aes(
    x = participation_ratio_delta,
    y = kappa_delta,
    color = filter_regime
  )
) +
  geom_point(size = 3) +
  facet_wrap(~ chip) +
  labs(
    title = "Complexity delta vs MP participation-ratio delta",
    x = "MP participation-ratio delta",
    y = "SVD kappa delta",
    color = "Filter regime"
  )
```

## Plot: Effective rank delta vs MP spectral entropy delta

```{r}
ggplot(
  combined_df,
  aes(
    x = spectral_entropy_delta,
    y = effrank_delta,
    color = filter_regime
  )
) +
  geom_point(size = 3) +
  facet_wrap(~ chip) +
  labs(
    title = "Effective rank delta vs MP spectral entropy delta",
    x = "MP spectral entropy delta",
    y = "Complexity effective-rank delta",
    color = "Filter regime"
  )
```

## Early interpretation

These joined diagnostics are central to the framework.

They ask whether classical complexity changes align with MP-derived spectral organization changes.

Potential outcomes:

| Pattern | Possible interpretation |
|---|---|
| `kappa_delta` aligns with MP concentration metrics | Complexity is tracking dominant spectral organization. |
| `effrank_delta` aligns with MP entropy/participation metrics | Effective rank is tracking distributed spectral organization. |
| Complexity and MP metrics diverge | The engines are capturing complementary structural dimensions. |

# Suggested next diagnostics

## Rank comparisons by largest complexity gain

```{r}
complexity_df |>
  dplyr::arrange(desc(kappa_delta)) |>
  dplyr::select(
    chip,
    group,
    comparison,
    filter_regime,
    kappa_delta,
    effrank_delta,
    composite_kappa_delta,
    p_perm,
    p_perm_fdr
  ) |>
  head(20)
```

## Rank comparisons by largest complexity loss

```{r}
complexity_df |>
  dplyr::arrange(kappa_delta) |>
  dplyr::select(
    chip,
    group,
    comparison,
    filter_regime,
    kappa_delta,
    effrank_delta,
    composite_kappa_delta,
    p_perm,
    p_perm_fdr
  ) |>
  head(20)
```

## Check regime robustness

```{r}
complexity_wide <- complexity_df |>
  dplyr::select(chip, group, comparison, filter_regime, kappa_delta, effrank_delta) |>
  tidyr::pivot_wider(
    names_from = filter_regime,
    values_from = c(kappa_delta, effrank_delta)
  )

dplyr::glimpse(complexity_wide)
```

# Cautions

The current complexity engine uses SVD-based condition-number metrics and related matrix summaries. These are sensitive to:

- sample size;
- probe count;
- rank deficiency;
- near-zero singular values;
- preprocessing and normalization;
- filtering regime;
- platform-specific probe design.

Therefore, current complexity outputs should be treated as structural descriptors and early exploratory inference rather than final biological claims.

The current run uses `gene_set_mode = "FULL"`. GO, KEGG, and MSigDB-stratified complexity analyses remain a later extension and should be added only after the FULL-level structural engine is stable.

# Working interpretation template

For each comparison, a preliminary interpretation can be generated from the direction of key complexity deltas:

| Pattern | Possible interpretation |
|---|---|
| `kappa_delta > 0` | Tumor matrix shows increased SVD-based condition structure. |
| `kappa_delta < 0` | Tumor matrix shows reduced SVD-based condition structure. |
| `effrank_delta > 0` | Tumor matrix shows broader effective dimensional participation. |
| `effrank_delta < 0` | Tumor matrix shows reduced effective dimensional participation. |
| `composite_kappa_delta > 0` | Multiple condition-number summaries suggest increased matrix conditioning. |
| `composite_kappa_delta < 0` | Multiple condition-number summaries suggest reduced matrix conditioning. |
| `kappa_delta > 0` and `effrank_delta > 0` | Increased complexity with broader effective-dimensional organization. |
| `kappa_delta > 0` and `effrank_delta < 0` | Increased condition structure concentrated in fewer effective dimensions. |

# Provisional conceptual summary

The complexity engine now provides the classical structural counterpart to the MP spectral engine.

The immediate purpose is to determine whether tumor-associated changes in matrix complexity align with MP-derived changes in spectral organization.

At this stage, the most important next questions are:

1. Are complexity deltas stable across `limma` and `variance_top3000`?
2. Are complexity deltas consistent across hu35ksuba and hu6800?
3. Does effective-rank change align with MP entropy and participation-ratio changes?
4. Does SVD kappa change align with MP dominant-eigenvalue or spectral-concentration changes?
5. Do particular cancer comparisons show consistent complexity gain or loss across chip and regime?

These results should be treated as provisional structural phenotypes to be tested against cancer type, platform, probe-selection regime, biological annotation, and later latent-space geometry.
