---
title: "Entropy Engine: Early Results Reference"
author: "Ali M. Al-Timimi"
date: "2026-05-09"
format:
  html:
    toc: true
    code-tools: true
  pdf:
    toc: true
execute:
  warning: true
  message: true
---

# Purpose

This document is a working reference for the first successful run of the entropy structural engine within the Global Cancer Complexity structural inference framework.

The goal is not to provide a final biological interpretation, but to preserve the current code, diagnostic plots, and early observations in a reproducible form.

At this stage, the entropy engine has successfully propagated the following identifiers across the structural inference pipeline:

- `chip`
- `group`
- `comparison`
- `filter_regime`
- `engine`
- `gene_set_mode`
- `gene_set_name`
- entropy metrics
- tumor-minus-normal delta metrics
- permutation-derived p-values
- FDR-adjusted p-values

The current implementation focuses on `gene_set_mode = "FULL"`, meaning entropy is computed over the full filtered probe set for each normal--tumor comparison and probe-selection regime.

# Conceptual framework

The entropy engine treats each normal--tumor comparison as a transformation in transcriptomic disorder and spectral organization.

Two complementary entropy measures are currently evaluated:

- **Shannon entropy:** expression-level uncertainty and distributional disorder;
- **Spectral entropy:** entropy of covariance-eigenvalue structure, reflecting the distribution of transcriptomic variance across effective modes.

Positive entropy deltas suggest increased transcriptomic disorder or
broader variance dispersion in tumor. Negative entropy deltas suggest increasing structural concentration, ordering, or reduced uncertainty in tumor.

The current focus is on whether tumors show evidence of:

- increased or decreased Shannon entropy;
- increased or decreased spectral entropy;
- concordance or divergence between Shannon and spectral entropy changes;
- consistency across chip platforms and probe-selection regimes;
- compatibility with complexity and Marchenko--Pastur spectral results.

## Why spectral entropy is central to the chaos/anti-chaos distinction

Although both Shannon entropy and spectral entropy are useful, spectral
entropy is currently the more direct structural measure for the
chaos/anti-chaos interpretation used in this framework.

Shannon entropy computed on raw expression values summarizes the
distributional uncertainty of expression magnitudes. It is therefore
sensitive to the marginal spread of values across the expression matrix.
This can be useful descriptively, but it does not directly distinguish
whether variation is organized across many coordinated modes or
concentrated into a smaller number of dominant modes.

Spectral entropy, by contrast, is computed from the covariance
eigenspectrum. It therefore summarizes how transcriptomic variance is
distributed across orthogonal modes of collective gene-expression
variation.

In this framework, a tumor transition is interpreted as more
**chaotic** when variance becomes more diffusely distributed across
multiple spectral modes. This corresponds to a broader eigenspectrum and
higher spectral entropy.

Conversely, a transition is interpreted as more **anti-chaotic** when
variance becomes more concentrated into fewer dominant modes. This
corresponds to spectral compression, lower spectral entropy, and a more
ordered covariance structure.

Thus:

| Pattern | Structural interpretation |
|---|---|
| `spectral_delta > 0` | Variance becomes more diffusely distributed across modes; interpreted as increased spectral disorder and potential chaotic tendency. |
| `spectral_delta < 0` | Variance becomes more concentrated into fewer modes; interpreted as increased spectral ordering or anti-chaotic tendency. |
| `spectral_delta ≈ 0` | No strong evidence for a major shift in covariance-spectrum organization. |

This distinction is important because cancer may alter not only the
amount of expression variability, but also the organization of that
variability. Spectral entropy is therefore closer to the structural
question motivating this project: whether tumorigenesis involves a gain,
loss, or reorganization of system-level transcriptomic order.

For this reason, Shannon entropy should currently be treated as a
supporting distributional descriptor, while spectral entropy should be
treated as the primary entropy-based structural descriptor.

## Spectral entropy as a proxy for chaotic tendency

The present framework does not claim to directly measure dynamical chaos
in the strict mathematical sense used in nonlinear dynamical systems
theory (e.g., Lyapunov exponents, strange attractors, or deterministic
chaotic trajectories).

Instead, spectral entropy is used as a structural proxy for the degree
to which transcriptomic variance is diffusely distributed across
multiple coordinated modes of system behavior.

Spectral entropy is computed from the covariance eigenspectrum. The
eigenvalues summarize how total transcriptomic variance is distributed
across orthogonal modes of collective variation.

When variance is concentrated into a small number of dominant modes, the
eigenspectrum is relatively ordered and spectrally compressed. This
produces lower spectral entropy and is interpreted here as a more
ordered or anti-chaotic structural state.

When variance is distributed more evenly across many modes, the
eigenspectrum becomes broader and less concentrated. This produces
higher spectral entropy and is interpreted as increased spectral
disorder or chaotic tendency.

The rationale for this interpretation is that highly organized systems
often exhibit constrained variance concentrated into relatively few
coordinated dimensions, whereas increasingly disordered or unstable
systems exhibit variance dispersion across larger numbers of effective
degrees of freedom.

Thus, spectral entropy is not treated here as a direct dynamical proof
of chaos, but rather as a systems-level descriptor of covariance
organization whose increase may reflect increasing structural disorder,
loss of coordinated regulation, or expansion of effective dynamical
degrees of freedom during tumorigenesis.

This interpretation also aligns naturally with related quantities used
elsewhere in the framework, including:

- effective rank;
- participation ratio;
- MP spectral entropy;
- latent-space anisotropy;
- covariance-spectrum dispersion.

Accordingly, spectral entropy is treated as a biologically interpretable
proxy for transcriptomic disorder and potential chaotic tendency rather
than as a literal dynamical-chaos measurement.

## Current entropy metrics

The main paired entropy output currently includes:

- `shannon_normal`
- `shannon_tumor`
- `shannon_delta`
- `shannon_direction`
- `spectral_normal`
- `spectral_tumor`
- `spectral_delta`
- `spectral_direction`
- `p_perm_spectral`
- `p_perm_shannon`
- `p_perm_shannon_fdr`

Depending on the current implementation of `compare_pair_entropy()`, the output may also include entropy bootstrap intervals. If these are not yet present, bootstrap inference should be treated as a planned extension rather than a current result.

### Statistical notes

These should be interpreted cautiously as structural descriptors rather than as finalized biological claims. Importantly, these metrics may capture partially distinct structural dimensions rather than redundant summaries of a single underlying quantity.

The permutation p-values (`p_perm_shannon`)/(`p_perm_spectral`)  are currently computed using a two-sided permutation framework:

```text
mean(abs(null_distribution) >= abs(observed_difference))
```

This choice reflects the current exploratory nature of the structural inference framework, where both entropy gain and entropy loss are considered biologically meaningful possibilities.

Adjusted permutation p-values (`p_perm_shannon_fdr`) are computed using the Benjamini--Hochberg procedure within each `chip × filter_regime × gene_set_mode` stratum.

Permutation procedures are seeded for reproducibility.

# Load results

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

source(
  here::here("R/config/global_cancer/structural_inference_config.R"),
  local = FALSE
)

entropy_results <- readRDS(
  file.path(structural_rdata_dir, "entropy_results.rds")
)
```

# Inspect object structure

```{r}
names(entropy_results)
str(entropy_results, max.level = 1)
```

Expected structure:

```text
summary: paired normal--tumor entropy summaries
```

# Inspect summary table

```{r}
entropy_df <- entropy_results$summary

dplyr::glimpse(entropy_df)
```

Check computation status:

```{r}
entropy_df |>
  dplyr::count(status)
```

Check that chip and regime propagation worked:

```{r}
entropy_df |>
  dplyr::count(chip, filter_regime)
```

Check unique probe-selection regimes:

```{r}
unique(entropy_df$filter_regime)
```

Expected regimes from current configuration:

```text
limma
variance_top3000
```

## Early interpretation

Successful recovery of both `limma` and `variance_top3000` confirms that the entropy engine now follows the same regime-aware structural architecture as the MP spectral and complexity engines.

This means the following chain is functioning:

```text
config
→ pipeline
→ wrapper
→ pairwise entropy engine
→ paired comparison layer
→ core entropy metrics
→ standardized summary table
```

The initial run may produce a small number of expected skips for limma comparisons with too few retained probes. These should be retained as QC information rather than treated as software failures.



# Plot 1: Shannon entropy delta by chip and regime

```{r}
ggplot(
  entropy_df,
  aes(
    x = filter_regime,
    y = shannon_delta
  )
) +
  geom_boxplot() +
  facet_wrap(~ chip) +
  labs(
    title = "Shannon entropy delta by chip and probe-selection regime",
    x = "Filter regime",
    y = "Tumor-minus-normal Shannon entropy delta"
  )
```

## Early interpretation

This plot provides a compact regime-level view of Shannon entropy gain or loss.

Positive `shannon_delta` indicates that tumor matrices have higher expression-level entropy than their matched normal matrices. Negative `shannon_delta` indicates reduced expression-level entropy in tumor.

The interpretation should remain cautious because Shannon entropy can be sensitive to the marginal distribution of expression values, preprocessing, feature count, and filtering regime.

# Plot 2: Spectral entropy delta by chip and regime

```{r}
ggplot(
  entropy_df,
  aes(
    x = filter_regime,
    y = spectral_delta
  )
) +
  geom_boxplot() +
  facet_wrap(~ chip) +
  labs(
    title = "Spectral entropy delta by chip and probe-selection regime",
    x = "Filter regime",
    y = "Tumor-minus-normal spectral entropy delta"
  )
```

## Early interpretation

Spectral entropy provides a complementary measure of how variance is distributed across covariance eigenmodes.

Positive `spectral_delta` suggests that tumor-associated variation is distributed more broadly across spectral modes. Negative `spectral_delta` suggests concentration of variance into fewer dominant modes.

This metric should later be compared directly with MP participation-ratio, MP spectral-entropy, effective-rank, and latent-space covariance metrics.

# Plot 3: Shannon delta vs spectral entropy delta

```{r}
ggplot(
  entropy_df,
  aes(
    x = shannon_delta,
    y = spectral_delta,
    color = filter_regime
  )
) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_point(size = 3) +
  facet_wrap(~ chip) +
  labs(
    title = "Shannon entropy delta vs spectral entropy delta",
    x = "Shannon entropy delta",
    y = "Spectral entropy delta",
    color = "Filter regime"
  )
```

## Early interpretation

This plot tests whether expression-level entropy and spectral entropy move together or diverge across tumor transitions.

| Pattern | Possible interpretation |
|---|---|
| `shannon_delta > 0`, `spectral_delta > 0` | Increased expression-level disorder with broader spectral variance dispersion. |
| `shannon_delta > 0`, `spectral_delta < 0` | Increased marginal uncertainty but more concentrated covariance structure. |
| `shannon_delta < 0`, `spectral_delta > 0` | Reduced marginal uncertainty but broader spectral organization. |
| `shannon_delta < 0`, `spectral_delta < 0` | Reduced entropy and increased spectral concentration. |

This distinction is central because Shannon and spectral entropy may represent different structural dimensions of tumor-associated transformation.

# Plot 4: Spectral entropy direction counts

```{r}
entropy_df |>
  dplyr::count(chip, filter_regime, spectral_direction) |>
  ggplot(
    aes(
      x = filter_regime,
      y = n,
      fill = spectral_direction
    )
  ) +
  geom_col(position = "dodge") +
  facet_wrap(~ chip) +
  labs(
    title = "Spectral entropy direction counts by chip and regime",
    x = "Filter regime",
    y = "Number of comparisons",
    fill = "Spectral direction"
  )
```

## Early interpretation

This plot provides a coarse summary of how often spectral entropy increases, decreases, or remains near neutral under each chip and probe-selection regime.

Spectral direction is emphasized here because it is structurally closer to the MP spectral engine and to effective-rank-style complexity measures than raw Shannon entropy.

This plot should not be treated as a final biological result, but it is useful for detecting gross asymmetry across platforms or filtering regimes.

# Plot 5: Shannon entropy direction counts

```{r}
entropy_df |>
  dplyr::count(chip, filter_regime, shannon_direction) |>
  ggplot(
    aes(
      x = filter_regime,
      y = n,
      fill = shannon_direction
    )
  ) +
  geom_col(position = "dodge") +
  facet_wrap(~ chip) +
  labs(
    title = "Shannon entropy direction counts by chip and regime",
    x = "Filter regime",
    y = "Number of comparisons",
    fill = "Shannon direction"
  )
```

## Early interpretation

This plot summarizes the expression-level entropy direction categories.

Comparing this plot with the spectral direction plot can help identify whether marginal expression disorder and covariance-spectral disorder are broadly concordant or whether they diverge by platform, regime, or tumor comparison.

# Correlation analysis of entropy metrics

```{r}
entropy_metrics <- entropy_df |>
  dplyr::select(
    shannon_delta,
    spectral_delta
  )

cor(entropy_metrics, use = "pairwise.complete.obs")
```

## Early interpretation

This correlation matrix tests whether Shannon and spectral entropy deltas behave as redundant summaries or as partially distinct descriptors.

A high positive correlation would suggest a shared entropy axis. A weak or negative correlation would suggest that marginal expression uncertainty and covariance-spectral organization capture different aspects of tumor-associated structural transformation.

# PCA of entropy delta metrics

```{r}
entropy_metrics_clean <- entropy_metrics |>
  tidyr::drop_na()

nonconstant_cols <- entropy_metrics_clean |>
  dplyr::summarise(
    dplyr::across(
      dplyr::everything(),
      ~ sd(.x, na.rm = TRUE) > 0
    )
  ) |>
  unlist()

entropy_metrics_pca <- entropy_metrics_clean[, nonconstant_cols, drop = FALSE]

if (ncol(entropy_metrics_pca) >= 2) {
  entropy_pca <- prcomp(
    entropy_metrics_pca,
    scale. = TRUE
  )
  summary(entropy_pca)
  round(entropy_pca$rotation, 3)
} else {
  message("PCA skipped: fewer than two nonconstant entropy delta metrics.")
}
```

## PCA score plot

```{r}
if (exists("entropy_pca")) {
  pca_row_idx <- stats::complete.cases(entropy_metrics) &
    apply(entropy_metrics, 1, function(x) all(is.finite(x)))
  
  entropy_pca_plot_df <- entropy_df[pca_row_idx, ] |>
    dplyr::select(chip, group, comparison, filter_regime) |>
    dplyr::bind_cols(as.data.frame(entropy_pca$x))
  
  ggplot(
    entropy_pca_plot_df,
    aes(
      x = PC1,
      y = PC2,
      color = filter_regime,
      shape = chip
    )
  ) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_vline(xintercept = 0, linetype = "dashed") +
    geom_point(size = 3) +
    labs(
      title = "PCA of entropy delta metrics",
      x = "PC1",
      y = "PC2",
      color = "Filter regime",
      shape = "Chip"
    )
}
```

## Early interpretation

With only Shannon and spectral entropy deltas, PCA is mainly a diagnostic convenience rather than a substantive dimensionality-reduction analysis.

If both entropy metrics are retained and nonconstant, PC1 may represent a shared entropy-gain/loss axis, while PC2 may represent divergence between expression-level entropy and covariance-spectral entropy.

This section becomes more useful if bootstrap summaries, additional entropy variants, or gene-set-level entropy modes are added later.

# Cross-engine alignment with complexity metrics

```{r}
complexity_path <- file.path(structural_rdata_dir, "complexity_results.rds")

if (file.exists(complexity_path)) {
  complexity_results <- readRDS(complexity_path)
  complexity_df <- complexity_results$summary
  
  entropy_complexity_df <- entropy_df |>
    dplyr::inner_join(
      complexity_df,
      by = c(
        "chip",
        "group",
        "comparison",
        "filter_regime",
        "gene_set_mode",
        "gene_set_name"
      ),
      suffix = c("_entropy", "_complexity")
    )
  
  dplyr::glimpse(entropy_complexity_df)
} else {
  message("Complexity results not found; skipping complexity alignment.")
}
```

## Plot: Spectral entropy delta vs effective-rank delta

```{r}
if (exists("entropy_complexity_df") && nrow(entropy_complexity_df) > 0) {
  ggplot(
    entropy_complexity_df,
    aes(
      x = effrank_delta,
      y = spectral_delta,
      color = filter_regime
    )
  ) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_vline(xintercept = 0, linetype = "dashed") +
    geom_point(size = 3) +
    facet_wrap(~ chip) +
    labs(
      title = "Spectral entropy delta vs effective-rank delta",
      x = "Effective-rank delta",
      y = "Spectral entropy delta",
      color = "Filter regime"
    )
}
```

## Plot: Shannon entropy delta vs SVD kappa delta

```{r}
if (exists("entropy_complexity_df") && nrow(entropy_complexity_df) > 0) {
  ggplot(
    entropy_complexity_df,
    aes(
      x = kappa_delta,
      y = shannon_delta,
      color = filter_regime
    )
  ) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_vline(xintercept = 0, linetype = "dashed") +
    geom_point(size = 3) +
    facet_wrap(~ chip) +
    labs(
      title = "Shannon entropy delta vs SVD kappa delta",
      x = "SVD kappa delta",
      y = "Shannon entropy delta",
      color = "Filter regime"
    )
}
```

## Early interpretation

This section begins to evaluate whether entropy shifts align with complexity shifts.

In particular, `spectral_delta` should be inspected alongside `effrank_delta`, because both summarize aspects of how variation is distributed across effective dimensions or eigenmodes.

By contrast, `shannon_delta` and `kappa_delta` may diverge because Shannon entropy is a distribution-level uncertainty measure whereas SVD kappa reflects matrix conditioning.

# Cross-engine alignment with MP spectral metrics

```{r}
mp_path <- file.path(structural_rdata_dir, "mp_results.rds")

if (file.exists(mp_path)) {
  mp_results <- readRDS(mp_path)
  mp_df <- mp_results$summary
  
  entropy_mp_df <- entropy_df |>
    dplyr::inner_join(
      mp_df,
      by = c(
        "chip",
        "group",
        "comparison",
        "filter_regime",
        "gene_set_mode",
        "gene_set_name"
      ),
      suffix = c("_entropy", "_mp")
    )
  
  dplyr::glimpse(entropy_mp_df)
} else {
  message("MP results not found; skipping MP alignment.")
}
```

## Early interpretation

This placeholder is included to preserve parallel structure with the complexity reference notebook.

The most relevant future comparisons are likely to include:

- entropy `spectral_delta` vs MP spectral entropy delta;
- entropy `spectral_delta` vs MP participation-ratio delta;
- entropy `shannon_delta` vs MP outlier or bulk summaries, if available.

Column names should be verified against the current MP output schema before adding plots here.

# Suggested next diagnostics

## Rank comparisons by largest Shannon entropy gain

```{r}
entropy_df |>
  dplyr::arrange(desc(shannon_delta)) |>
  dplyr::select(
    chip,
    filter_regime,
    group,
    comparison,
    shannon_normal,
    shannon_tumor,
    shannon_delta,
    shannon_direction,
    p_perm_shannon,
    p_perm_shannon_fdr
  ) |>
  head(15)
```

## Rank comparisons by largest Shannon entropy loss

```{r}
entropy_df |>
  dplyr::arrange(shannon_delta) |>
  dplyr::select(
    chip,
    filter_regime,
    group,
    comparison,
    shannon_normal,
    shannon_tumor,
    shannon_delta,
    shannon_direction,
    p_perm_shannon,
    p_perm_shannon_fdr
  ) |>
  head(15)
```

## Rank comparisons by largest spectral entropy gain

```{r}
entropy_df |>
  dplyr::arrange(desc(spectral_delta)) |>
  dplyr::select(
    chip,
    filter_regime,
    group,
    comparison,
    spectral_normal,
    spectral_tumor,
    spectral_delta,
    spectral_direction,
    p_perm_spectral
  ) |>
  head(15)
```

## Rank comparisons by largest spectral entropy loss

```{r}
entropy_df |>
  dplyr::arrange(spectral_delta) |>
  dplyr::select(
    chip,
    filter_regime,
    group,
    comparison,
    spectral_normal,
    spectral_tumor,
    spectral_delta,
    spectral_direction,
    p_perm_spectral
  ) |>
  head(15)
```

## Check regime robustness

```{r}
entropy_df |>
  dplyr::select(
    chip,
    group,
    comparison,
    filter_regime,
    shannon_delta,
    spectral_delta
  ) |>
  tidyr::pivot_wider(
    names_from = filter_regime,
    values_from = c(shannon_delta, spectral_delta)
  )
```

# Cautions

The entropy engine is now structurally integrated, but the current results should remain exploratory.

Important cautions:

- Shannon entropy and spectral entropy are not interchangeable.
- Entropy changes should not be interpreted as direct evidence of biological chaos without additional support.
- Probe-selection regime, sample size, matrix rank, and normalization choices may influence entropy summaries.
- Limma-based regimes can produce skipped comparisons when fewer than five selected probes are retained.
- FDR-adjusted permutation results should be interpreted within the defined `chip × filter_regime × gene_set_mode` stratum.
- Cross-engine interpretation should wait until entropy, complexity, MP, and latent-space summaries are compared under matched identifiers.

# Working interpretation template

For each comparison, a cautious preliminary interpretation can use the following structure:

```text
For [comparison] on [chip] under [filter_regime], tumor samples showed
[an increase / a decrease / little change] in Shannon entropy and
[an increase / a decrease / little change] in spectral entropy relative
to matched normal samples.

This suggests a possible shift in [expression-level uncertainty / spectral
variance organization / both], but this should be interpreted as a structural
descriptor rather than a finalized biological claim.

Permutation evidence was [weak / moderate / strong] before FDR correction
and [retained / not retained] after FDR adjustment within the relevant
chip × filter_regime × gene_set_mode stratum.
```

# Provisional conceptual summary

The entropy structural engine now provides a regime-aware, chip-aware, comparison-aware summary of tumor-minus-normal entropy shifts.

The central interpretive distinction is between:

- **Shannon entropy**, which summarizes expression-level uncertainty; and
- **spectral entropy**, which summarizes the distribution of covariance structure across eigenmodes.

The main working hypothesis is not simply that cancer gains or loses entropy globally, but that different tumor transitions may show different combinations of marginal disorder and spectral reorganization.

These entropy signatures should ultimately be interpreted alongside:

- complexity deltas;
- MP spectral metrics;
- latent-space geometry;
- GO/KEGG/MSigDB biological annotations;
- comparison-specific normal and tumor sample composition.
