---
title: "Marchenko--Pastur Spectral Engine: Early Results Reference"
author: "Ali M. Al-Timimi"
date: "2026-05-08"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
  pdf:
    toc: true
execute:
  echo: true
  warning: true
  message: true
---

# Purpose

This document is a working reference for the first successful run of the Marchenko--Pastur (MP) spectral engine within the Global Cancer Complexity structural inference framework.

The goal is not to provide a final interpretation, but to preserve the current code, diagnostic plots, and early observations in a reproducible form.

At this stage, the MP engine has successfully propagated the following identifiers across the structural inference pipeline:

- `chip`
- `group`
- `comparison`
- `filter_regime`
- `condition`
- MP/spectral summary metrics
- tumor-minus-normal delta metrics

# Conceptual framing

The MP spectral engine treats each normal--tumor comparison as a transformation in the eigenvalue spectrum of a probe-defined expression space.

For each comparison, the engine computes condition-level spectral summaries for normal and tumor matrices, then derives tumor-minus-normal deltas.

The current focus is on whether tumors show evidence of:

- increased or decreased spectral concentration;
- redistribution of eigenvalue mass;
- increased or decreased effective dimensionality;
- emergence of additional super-MP spectral modes;
- consistency across chip platforms and probe-selection regimes.

# Load results

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

source(here::here("R/config/global_cancer/structural_inference_config.R"), local = FALSE)

mp_spectral_results <- readRDS(
  file.path(structural_rdata_dir, "mp_spectral_results.rds")
)
```

# Inspect object structure

```{r}
names(mp_spectral_results)
str(mp_spectral_results, max.level = 1)
```

Expected structure:

```text
summary: condition-level MP/spectral summaries
deltas : tumor-minus-normal spectral deltas
```

# Inspect condition-level summaries

```{r}
summary_df <- mp_spectral_results$summary

dplyr::glimpse(summary_df)
```

Check computation status:

```{r}
table(summary_df$status)
```

Early observed result from the first successful run:

```text
ok                             102
skipped_insufficient_dimensions  4
```

## Early interpretation

Most condition-level spectral summaries were computed successfully. A small number of skipped cases are expected when one side of a comparison has too few samples or too few retained probes after probe-selection and matrix intersection.

These skipped cases should be preserved as QC outcomes rather than silently removed.

# Inspect delta table

```{r}
delta_df <- mp_spectral_results$deltas

dplyr::glimpse(delta_df)
```

Check that chip and regime propagation worked:

```{r}
delta_df |>
  dplyr::count(chip, filter_regime)
```

Check unique probe-selection regimes:

```{r}
unique(delta_df$filter_regime)
```

Expected regimes from current configuration:

```text
limma
variance_top3000
```

## Early interpretation

Successful recovery of both `limma` and `variance_top3000` confirms that the refactored naming chain is functioning across:

```text
config
→ pipeline
→ wrapper
→ pairwise spectral engine
→ paired comparison layer
→ single-condition spectral layer
→ MP helper layer
→ delta table
```

This is an important software milestone because MP outputs are now regime-aware.

# Basic MP delta metrics

The main tumor-minus-normal delta metrics currently include:

- `largest_eigenvalue_delta`
- `largest_eigenvalue_fraction_delta`
- `n_spikes_delta`
- `spike_fraction_delta`
- `excess_spectral_mass_delta`
- `spectral_entropy_delta`
- `participation_ratio_delta`

These should be interpreted cautiously as structural descriptors, not yet as final biological claims.

# Plot 1: Participation ratio delta vs spectral entropy delta

```{r}
ggplot(
  delta_df,
  aes(
    x = participation_ratio_delta,
    y = spectral_entropy_delta,
    color = filter_regime
  )
) +
  geom_point(size = 3) +
  facet_wrap(~ chip) +
  labs(
    title = "Participation ratio delta vs spectral entropy delta",
    x = "Participation ratio delta",
    y = "Spectral entropy delta",
    color = "Filter regime"
  )
```

## Early interpretation

The first successful run showed a strong positive association between `participation_ratio_delta` and `spectral_entropy_delta`.

This is structurally coherent: as eigenvalue mass becomes more evenly distributed, spectral entropy increases and more effective eigenmodes contribute to the spectrum.

This relationship suggests that these metrics are not behaving as unrelated descriptors. Instead, they appear to track a common organizational transformation.

# Plot 2: Spectral entropy delta by chip and regime

```{r}
ggplot(
  delta_df,
  aes(
    x = filter_regime,
    y = spectral_entropy_delta
  )
) +
  geom_boxplot() +
  facet_wrap(~ chip) +
  labs(
    title = "Spectral entropy delta by chip and probe-selection regime",
    x = "Filter regime",
    y = "Spectral entropy delta"
  )
```

## Early interpretation

The boxplots provide a compact regime-level view of tumor-minus-normal spectral entropy changes.

The early results suggest that many tumor comparisons show positive entropy shifts, but the effect is not uniform. Some comparisons show negative or near-zero deltas.

This supports a more nuanced interpretation than “cancer equals disorder.” The framework may instead distinguish different kinds of spectral reorganization.

# Plot 3: Participation ratio delta vs dominant eigenvalue fraction delta

```{r}
ggplot(
  delta_df,
  aes(
    x = participation_ratio_delta,
    y = largest_eigenvalue_fraction_delta,
    color = filter_regime
  )
) +
  geom_point(size = 3) +
  facet_wrap(~ chip) +
  labs(
    title = "Participation ratio delta vs largest eigenvalue fraction delta",
    x = "Participation ratio delta",
    y = "Largest eigenvalue fraction delta",
    color = "Filter regime"
  )
```

## Early interpretation

This plot showed a strong inverse relationship between `participation_ratio_delta` and `largest_eigenvalue_fraction_delta`.

Interpretively, this suggests that when the leading eigenmode loses relative dominance, spectral organization becomes more distributed across multiple modes.

This relationship is important because it links three core concepts:

- dominant-mode concentration;
- effective dimensionality;
- distributed spectral organization.

# Plot 4: Spike count delta vs spectral entropy delta

```{r}
ggplot(
  delta_df,
  aes(
    x = n_spikes_delta,
    y = spectral_entropy_delta,
    color = filter_regime
  )
) +
  geom_point(size = 3) +
  facet_wrap(~ chip) +
  labs(
    title = "MP spike count delta vs spectral entropy delta",
    x = "MP spike count delta",
    y = "Spectral entropy delta",
    color = "Filter regime"
  )
```

## Early interpretation

The early result suggests that increased spike count is often associated with increased spectral entropy.

This is potentially important because additional super-MP modes do not necessarily imply collapse into a single dominant program. In these results, additional spikes often accompany broader spectral participation.

This may indicate structured diversification of spectral organization in some tumor transitions.

# Correlation analysis of delta metrics

```{r}
delta_metrics <- delta_df |>
  dplyr::select(
    largest_eigenvalue_delta,
    largest_eigenvalue_fraction_delta,
    n_spikes_delta,
    spike_fraction_delta,
    excess_spectral_mass_delta,
    spectral_entropy_delta,
    participation_ratio_delta
  )

cor(delta_metrics, use = "pairwise.complete.obs")
```

## Early observed correlations

Key correlations from the first successful run:

```text
participation_ratio_delta vs spectral_entropy_delta      ≈  0.93
participation_ratio_delta vs largest_eigenvalue_fraction_delta ≈ -0.96
spectral_entropy_delta vs largest_eigenvalue_fraction_delta    ≈ -0.92
largest_eigenvalue_delta vs excess_spectral_mass_delta         ≈  0.83
n_spikes_delta vs participation_ratio_delta                    ≈  0.71
```

## Early interpretation

The correlation matrix suggests that the MP-derived metrics are organized around a dominant structural axis.

One pole appears to represent distributed spectral organization:

- increased spectral entropy;
- increased participation ratio;
- increased spike count;
- reduced leading eigenvalue fraction.

The opposite pole appears to represent concentrated spectral organization:

- increased leading eigenvalue dominance;
- increased excess spectral mass;
- reduced entropy;
- reduced effective dimensionality.

This may become a useful distinction between distributed restructuring and concentrated restructuring.

# PCA of MP delta metrics

```{r}
pca_res <- prcomp(
  na.omit(delta_metrics),
  scale. = TRUE
)

summary(pca_res)
```

Inspect loadings:

```{r}
round(pca_res$rotation, 3)
```

## Early observed PCA result

The first run gave approximately:

```text
PC1 proportion of variance: 0.7366
PC2 proportion of variance: 0.1362
Cumulative PC1 + PC2:       0.8728
```

## Early interpretation

The fact that PC1 explains approximately 74% of the variance suggests that these MP delta metrics largely collapse onto a dominant spectral-organization axis.

This is potentially significant. It suggests that tumor-associated spectral changes may not be a collection of unrelated metric shifts, but rather a coordinated transformation in spectral organization.

PC2 may represent a secondary axis, possibly separating spike structure, excess spectral mass, or chip/regime-specific behavior from the main distributed-versus-concentrated spectral axis. This requires inspection of PCA loadings and later biological annotation.

# Suggested next diagnostics

## PCA score plot by chip and regime

```{r}
pca_scores <- as.data.frame(pca_res$x) |>
  dplyr::mutate(
    row_id = dplyr::row_number()
  )

pca_metadata <- delta_df |>
  dplyr::select(chip, group, comparison, filter_regime) |>
  tidyr::drop_na()

pca_plot_df <- dplyr::bind_cols(
  pca_metadata[seq_len(nrow(pca_scores)), ],
  pca_scores
)

ggplot(
  pca_plot_df,
  aes(
    x = PC1,
    y = PC2,
    color = filter_regime,
    shape = chip
  )
) +
  geom_point(size = 3) +
  labs(
    title = "PCA of MP spectral delta metrics",
    x = "PC1",
    y = "PC2",
    color = "Filter regime",
    shape = "Chip"
  )
```

Note: if rows with missing values are removed before PCA, metadata alignment should be handled carefully. A safer version is provided below.

## Safer PCA score plot with explicit complete-case alignment

```{r}
complete_idx <- stats::complete.cases(delta_metrics)

pca_res_safe <- prcomp(
  delta_metrics[complete_idx, ],
  scale. = TRUE
)

pca_plot_df_safe <- delta_df[complete_idx, ] |>
  dplyr::select(chip, group, comparison, filter_regime) |>
  dplyr::bind_cols(as.data.frame(pca_res_safe$x))

ggplot(
  pca_plot_df_safe,
  aes(
    x = PC1,
    y = PC2,
    color = filter_regime,
    shape = chip
  )
) +
  geom_point(size = 3) +
  labs(
    title = "PCA of MP spectral delta metrics",
    x = "PC1",
    y = "PC2",
    color = "Filter regime",
    shape = "Chip"
  )
```

# Cautions

The current MP implementation should be interpreted carefully because many comparisons have very large feature-to-sample ratios.

For example, `q = n_features / n_samples` can be very large. Classical MP theory is most stable under appropriate random-matrix asymptotics, and small sample sizes may distort the empirical spectrum.

Therefore, current MP outputs should be treated as structural spectral descriptors and exploratory inference rather than final strict random-matrix-theoretic claims.

Potential future refinements include:

- shrinkage covariance/correlation estimation;
- empirical null simulation;
- randomized spectral diagnostics;
- sensitivity analysis across `min_samples`;
- comparison of correlation-based vs covariance-based spectra;
- chip-specific robustness analysis;
- biological annotation of extreme PC1/PC2 comparisons.

# Working interpretation template

For each comparison, a preliminary interpretation can be generated from the direction of key MP deltas:

| Pattern | Possible interpretation |
|---|---|
| `spectral_entropy_delta > 0` and `participation_ratio_delta > 0` | Tumor spectrum becomes more distributed across eigenmodes. |
| `largest_eigenvalue_fraction_delta < 0` | Dominant spectral mode loses relative concentration. |
| `n_spikes_delta > 0` | More super-MP structured modes emerge. |
| `excess_spectral_mass_delta > 0` | Structured spectral mass above the MP edge increases. |
| `spectral_entropy_delta < 0` and `participation_ratio_delta < 0` | Tumor spectrum becomes more concentrated. |
| `largest_eigenvalue_fraction_delta > 0` | Leading eigenmode becomes more dominant. |

# Provisional conceptual summary

The early MP results support a distinction between at least two spectral reorganization patterns:

## Distributed restructuring

Characterized by:

- increased spectral entropy;
- increased participation ratio;
- reduced largest eigenvalue fraction;
- often increased spike count.

This suggests that tumor-associated organization is redistributed across multiple spectral modes.

## Concentrated restructuring

Characterized by:

- increased largest eigenvalue or largest eigenvalue fraction;
- increased excess spectral mass;
- reduced spectral entropy;
- reduced participation ratio.

This suggests that tumor-associated organization may collapse into fewer dominant coordinated modes.

These patterns are not yet final biological categories. They are provisional structural phenotypes to be tested against cancer type, chip platform, probe-selection regime, GO/KEGG interpretation, and later latent-space geometry.
